\chapter{Introduction}

You are reading the main documentation of the HelUSB3 project. The main goal of
the project was to extend the existing USB support in HelenOS with support for
USB 3, namely implementing the driver for host controllers compliant with the
eXtensible Host Controller Interface (xHCI), extending the USB device driver
framework to USB work at USB 3 speeds, xHCI-only devices and overall refactoring and cleanup of
the USB subsystem in HelenOS. The previous support for USB was implemented as
a software project back in 2011, then was maintained and extended by HelenOS
developers (mostly by one team member of the project).

\section{Structure of This Document}

In the first chapter, we try to build context needed to understand the
following chapters. You can get an idea of what HelenOS is and how drivers work
there. Then, we try to introduce the reader to idioms and paradigms used in the
USB world, as the rest of the documentation assumes the reader has this basic
knowledge. All of this information is included only for the sake of
completeness, we do not aim to replace (nor duplicate) the documentation of
HelenOS nor USB specifications. Please consider these sections as informative
and supportive only, and follow on topics in their respective documentations.

In the second chapter, we focus on the xHCI driver, which is a major part of
the project. We describe its overall architecture and discuss implementation
decisions. This chapter, together with in-code documentation and knowledge of the
xHCI specification, shall provide reader enough information to modify and
extend our implementation without much trouble.

The third chapter is dedicated to modifications to the existing USB subsystem.
We do not discuss the obvious ones (like support for USB 3 hubs), but those
that would be avoidable if we took the assignment to the word. Since we like to
leave the table clean, we had to change some design decisions of the previous
authors. Some of the changes introduce more systematic approach, some were
needed to implement additional parts of the assignment without using a duct
tape. Some of them are just optimizations towards better performance or
usability. Also, the documentation of the previous project was obsoletes by
recent developments in HelenOS, so we try to document the current state of
the USB subsystem.

To support the implementation process and to allow us to test it in
a semi-automatic way, we have developed a standalone testing mechanism that
tests the USB stack end-to-end by using modified virtual environment. This is
the topic of Chapter~\ref{appendix:testing}.

The last chapter concludes the documentation with a brief summary of the
accomplished project goals, instructions on how to access the materials stored
online and on the attached CD. In addition, it outlines possible future
development of the USB framework following the project end.

As the USB subsystem is mostly useless without drivers for actual USB devices,
we would like to lessen the initial barrier that stands in the way of
implementing new USB device drivers. The Appendix~\ref{chap:usb-drivers} is
written as a complete guide to writing device drivers, hiding implementation
details of the stack and focusing on how to use it only. It still requires the
reader to understand the USB architecture and mechanics though, as well as
understanding the device they are writing the driver for.

The Appendix~\ref{sec:spec} includes the official project specification (in
Czech) and a concise timeline of the project realization throughout the years
2017 and 2018.

\section{About HelenOS}
HelenOS is a microkernel operating system. Contrary to operating systems with
monolithic kernels, key operating system functionality (including device
drivers, filesystems and networking) is implemented in user-space processes (or
servers) that interact via message passing interface. The rationale of this
decision is that the failure of one component, e.g.\ a faulty driver, does not
result in the crash of the entire system. It also allows a more modular system
design.

HelenOS was started in 2004 as a software project at the Faculty of Mathematics
and Physics at Charles University and currently is being developed by students,
former students and staff of this university, among with a number of
contributors around the world. It has participated in Google Summer of Code
several times. HelenOS is used as a research operating system and as a platform
for bachelor and master theses and student software projects.

HelenOS supports several architectures (among them ia32, amd64, 32-bit ARM and
big endian MIPS) and is released under the BSD license.

\subsection{xHCI Support in Operating Systems}
\label{subsect:support-in-oses}

Extensible Host Controller Interface was first drafted in 2008 and the final
version of the specification was released in 2010. Since then, xHCI support
arrived in all mainstream operating systems like Windows, Linux, BSDs and Mac
OS X.

Regarding special-purpose, microkernel and research OSes, xHCI support is not
yet widespread.  Most notably, xHCI support is included in Google's
microkernel-based Fuchsia and in bare-metal iPXE.  Haiku Project is currently
developing xHCI driver for their OS.

In HelenOS, the support for USB was started by the HelUSB project defended in 2011. In that
time, the USB driver framework was designed, and delivered with a few USB
drivers. From the Host Controller side, UHCI and OHCI were supported almost
completely. As for the device drivers, only a generic HID driver was provided
to demonstrate the functionality of the framework. Since the project was
delivered, the USB stack evolved a little and new drivers were implemented.
The EHCI support was rather rudimentary and there was no support for xHCI.

\section{Drivers in HelenOS}

Drivers in HelenOS are separate userspace tasks. There is a common framework
for writing device drivers, the \emph{Device Driver Framework}, abbreviated as
DDF. The DDF consists of two major parts: the \texttt{devman} service taking
care about starting drivers, attaching newly found devices to drivers, and many
more; and a \lib{libdrv} library creating a platform to write DDF-compatible
drivers on.

A driver, once started, is essentially a server task. It fills a callback table
with driver operations, and waits for being called by the library. Once the
devman wants the driver to take care of a new \emph{device}, its
\fnc{dev_add} callback is called. The driver is then supposed to do whatever
it needs to make the device operating, eventually creating \emph{functions} --
nodes that are passed back to devman to be taken care of. The function can be
either an \emph{exposed} function, being a leaf node usable directly by the user
(keyboard, disk partition, \dots), or an \emph{inner} function, essentially
being a \textit{device} for another driver.

Drivers then communicate with each other using \emph{interfaces}. Every
function is provided with a set of interfaces it provides to the child driver. The
driver itself then serves as a mediator or translator of operations called on
its functions to yet another operations called on its device. For example, the
\emph{PCI} driver exposes a function for every card physically attached to the
bus. One of its functions is an xHC, which is given as a device to the
\emph{xHCI} driver. It then creates a function for every gadget connected
through USB. USB functions are provided with an interface to issue USB
transfers. USB device drivers use this interface to actually drive the USB
devices. The xHCI driver translates the transfer requests to writes into PIO
space offered by the PCI driver as a part of the PCI function interface.

In the very end, the device tree exposes only its leaves. Every leaf node
(exposed function) can be added into several categories. Categories being
a well-known classes of devices (e.g.\ a disk partition) provide an interface
of a service. For the user, the whole DDF ecosystem looks like a modular
implementation of the service interface, allowing them, for example, to mount
the partition regardless of where it is located and how it is connected to the
system.

\input{01_usb}

\section{Existing USB Subsystem}

As mentioned in subsection~\ref{subsect:support-in-oses}, HelenOS already supported
UHCI, OHCI and to some extent EHCI.

The original documentation of the HelUSB project described the state of affairs
as the project ended. Since then, several improvements in both USB support
and in HelenOS driver framework in general were made, for example new USB
drivers like mass-storage were added. Unfortunately, these changes are pretty
much not documented anywhere. In this section, we will focus on the things that changed since the
documentation was written. We do not aim to replace the original documentation
of HelUSB project.

\vspace{1.5em}
\textsl{By all means, information in this chapter is written with
regard to the state before \emph{our} project was implemented. Thus, great part
of the information given in this section is already obsolete, but it's needed to
assess the damage we're personally responsible for.}
\vspace{1.5em}

The USB framework defines two classes of drivers -- the host controller drivers
and USB function drivers. For the first class, there is a library called
\lib{libusbhost} that aids in providing the unified interface of the host
controller to USB function drivers, and also implements common HC
functionality. There are four HC drivers at the moment:

\begin{itemize}
\item
	\textbf{VHC}, Virtual Host Controller. Implemented in the early phase of
	HelUSB project, served probably as a dummy backend to allow better work
	parallelization and debugging.

\item
	\textbf{UHCI}, Universal Host Controller Interface driver. The earliest
	interface supporting speeds of USB 1.0: Low- and Full-speed devices.
	Important for running HelenOS under QEMU, as it's the interface of the
	default HC QEMU emulates. Apart from isochronous transfers, the driver
	covers all functionality UHCI provides.

\item
	\textbf{OHCI}, Open Host Controller Interface driver. Somewhat complete,
	yet a bit simplified, especially in terms of transfer scheduling. Does not
	care about the polling interval, but schedules all interrupt transfers on
	every frame. Isochronous transfers not supported.

\item
	\textbf{EHCI}, Enhanced Host Controller Interface driver. Mostly a copy of
	OHCI driver, as it uses similar structures. Shall support USB 2 speeds, but
	the support is very limited -- the driver cannot use High-speed hubs to
	communicate with Full- and Low-speed devices, as the support for
	Transaction Translation is completely broken. Also, the bandwidth
	accounting is not implemented for High speed. Neither this controller
	supports isochronous transfers.
\end{itemize}

The HC driver is no longer split in half (as the project documentation states),
but all HC drivers emulate a virtual hub that is driven by a regular
\lib{usbhub} driver. The tree physical topology of USB is kept only inside the
HC driver, and is presented flat to the Device Driver Framework -- all USB
devices are child functions of the HC driver. They communicate with each other
through the DDF driver interface called \texttt{usb\_iface}, which contains all
methods various drivers use.

When the driver enumerates a USB function, it is usually taken care of by the
\lib{usbmid} driver. This driver scans the device descriptor for provided
interfaces, and creates child functions for them. These functions are then
driven by class drivers. Notable exception being the \lib{usbhub} driver,
taking care of the device directly, as hubs are not allowed to have multiple
interfaces.

The USB function drivers are well supported by the \lib{libusbdev} library.
This library builds an abstraction layer above \lib{libdrv}, used by other
drivers directly, to better suit the needs of USB devices. It does a complete
initialization of the USB device, including initiating a separate IPC
connection to the HC driver directly -- to avoid bouncing all operations in the
\lib{usbmid} driver. For this purpose, the \texttt{usb\_iface} contains two
methods: \fnc{get_my_iface} and \fnc{get_my_interface_handle}. The first one is
answered by the \lib{usbmid} driver with the number of interface driven, or
with a value of $-1$ by the HC driver if the driver serves the whole device.
The \fnc{get_my_interface_handle} call is recursive, until it reaches the HC
driver -- which answers it with devman handle of the USB device function. The
device driver then uses it to initiate connection to the HC driver, the same
way as the usbmid driver does.

Although this is sort-of hacky solution (the devman handle is supposed to be
private), it is currently the only one. Ideally, the drivers would use
a special method to let new connection forward to the HC driver, but for
complicated reasons, this does not work as expected. We discussed this with the
current HelenOS developers, and they confirmed us that the issue is still not
solved.

The \lib{usbhub} driver uses another four methods defined by the
\texttt{usb\_iface}. The interface methods \fnc{reserve_default_address} and
\fnc{release_default_address} ensure synchronization across multiple hubs
(possibly across multiple hub drivers), as the software must ensure that only one
device is listening on the default address at the same time. Then,
\fnc{device_enumerate} and \fnc{device_remove} announce that a device is
connected to (detached from) the hub, to be enumerated (removed) by the HC
driver. The interesting part is that the hub driver has no access to the
created device, as the logical topology presented to the DDF is flat.

All USB device drivers specify the endpoints they expect from the device in
a form of a static description, which they pass to the \lib{libusbdev} library
during driver initialization. Once a new device is added, the library
fetches the device descriptor and matches available endpoints against the
specification provided by the driver. Then the library creates \emph{pipes} --
an abstraction of endpoints based on their properties, not their exact numbers,
which are usually implementation defined. Pipes are then used by the driver
as, well, pipes to push data through and read data from.

The pipe creation process and their usage define the last four methods of which
the \texttt{usb\_iface} is comprised of: \fnc{(un)register_endpoint}, \fnc{read}
and \fnc{write}. The endpoint (un)registration informs the HC driver about
a pipe creation/disposal, and \fnc{read}/\fnc{write} methods are used to
actually transmit packets. Note that the interface is unified regardless of the
transfer type used by the endpoint.

As for the drivers available, there is a solid support for USB HID devices,
implementing keyboards, mice and multimedia keys. Also, a driver for USB Mass
storage exists and somehow works, despite several warnings and errors printed
to the log. Also, a fallback driver is provided to handle any USB device, to
enable the device examination for devices without their own driver, mainly for
debugging purposes.

Not to forget, there are two userspace utilities related to the USB stack. One
of them, \texttt{mkbd}, is not so important, as it is used only to demonstrate
functionality of multimedia keys HID driver. The other one, \texttt{usbinfo},
can be used to list available USB devices:

\begin{bdsh}
/ # usbinfo -l
Bus 37: /hw/pci0/00:04.0/ctl
	Device 61: /hw/pci0/00:04.0/usb1-fs
	Device 65: /hw/pci0/00:04.0/usb2-ls
\end{bdsh}

Other use-cases for this utility include descriptor dumping and examination of
device status.

It needs to be said that whole structure of the USB framework (and also the
DDF in general) expects the drivers to behave correctly and does not
implement any countermeasures against malicious behaviour of drivers. For
example, the \texttt{usbinfo} utility connects directly to the HC in the same
way as the device driver does, and fetches the device descriptor. In fact, any
other task can communicate directly with any USB device. Or, any driver can
call the interface methods designed for hubs only -- for example, it can
reserve the default address and never release it. Due to the experimental and
in-development nature of HelenOS, this is not an important problem. Yet, it is
an obstacle to solve before HelenOS will be ready for ``normal'' users, and it
will be a tough one.

Another thing related to the whole USB stack is that support for device
removal is in fact non-existent. At the time of HelUSB project, there was no
support for device removal in the Device Driver Framework, so it's not
surprising that the USB framework inherited this. There are attempts to
terminate interaction and release resources in case of repeated communication
errors.
